---
title: "chemo-dd: data organisation"
output: html_notebook
author: James Orr
---

Organising the data (colony counts, od reads, outflow volumes) from the density-dependence experiments with e. coli in chemostats in the Letten lab. 

dd2, dd3, and dd4 conducted by Kaleigh Davis

dd6, and dd7 conducted by James Orr and Alicia Williams


### Set up environment 

Load packages and clear environment 

```{r}
#### Required packages
library(tidyverse)        

#### Clear  environment 
rm(list = ls())   
```

Load data

```{r}
# cfu data
cfu_jo <- read.csv("data/cfu/cfu_counts_jo.csv")
cfu_kd <- read.csv("data/cfu/cfu_counts_kd.csv")

# od data
od_jo <- read.csv("data/od/od_jo.csv")
od_kd <- read.csv("data/od/od_kd.csv")

# outflow data
outflow_jo <- read.csv("data/outflow/outflow_rates_jo.csv")
outflow_kd <- read.csv("data/outflow/outflow_rates_kd.csv")
```

### Colony counts

Merge data and do some initial filtering 

```{r}
# join two dataframes
cfu <- cfu_jo %>%
  select(-inflow_setting)  %>%
  bind_rows(cfu_kd) %>%
  select(-date)

# filtering out data that we can't use
cfu <- cfu %>%
  # remove the sample that was taken after inflow pumps in dd6
  filter(time != "a") %>%
  select(-c(time)) %>%
  # remove overnights
  filter(day != "0") %>%
  # remove media tests
  filter(sample_media != "media") %>%
  select(-c(sample_media)) %>%
  # keep on lb plates
  filter(plate_type == "lb") %>%
  select(-c(plate_type))

# remove old dataframes
rm(list = c("cfu_jo", "cfu_kd"))   

```

Average counts

```{r}
cfu <- cfu %>%
  mutate(av_counts = rowMeans(across(dot1:dot10), na.rm = TRUE))
```

Densities

```{r}
cfu <- cfu %>%
  
  # multiply by 10^dilution factor to get CFUs per ul 
  # first row is 100ul of sample so it would be: count * 10 ^ 1 to get 1ul
  # six row would be: count * 10 ^ 6 to get to 1ul
  # remember that the very first row is diluted!
  mutate(cfu_undil = av_counts * 10^dilution_factor) %>%
  
  # each dot is 5ul, so divide by 5 to get counts per ul
  mutate(cfu_ul = cfu_undil/5)
```

Final dataframe

```{r}
cfu <- cfu %>%
  select(c(experiment_replicate, day, ID, cfu_ul))
```

### Outflow data

Merge dataframes

```{r}
outflow <- outflow_jo %>%
  # get the names the same
  rename(volume_out_round = volume_out) %>%
  rename(time_elapsed_round = time_elapsed) %>%
  # merge dataframes
  bind_rows(outflow_kd) %>%
  # remove some columns 
  select(-c(volume_out_full, time_elapsed_full, 
            exp_start_date, day_measured, species,
            media.bottle)) %>%
  # remove dd1 (failed after 10 hours)
  filter(experiment_rep != "dd1")

# remove old dataframes
rm(list = c("outflow_jo", "outflow_kd"))   
```

Get volume per hour 

```{r}
outflow <- outflow %>%
  mutate(outflow_per_hour = volume_out_round/time_elapsed_round) 
```

Average across each measurement of the same reactor 

```{r}
outflow <- outflow %>%
  group_by(bioreactor, experiment_rep, inflow_setting) %>%
  summarise(outflow_per_hour_mean = mean(outflow_per_hour))

#plot(outflow$inflow_setting, outflow$outflow_per_hour_mean)
```

Calculate dilution rate and final cleaning 

```{r}
outflow <- outflow %>%
  
  # as a fraction of the volume of the vial (21ml) to get rate of dilution 
  mutate(dilution_rate = outflow_per_hour_mean/21) %>%
  
  # rename some variables to match cfu and od
  rename(experiment_replicate = experiment_rep) %>%
  rename(ID = bioreactor)
```



### OD data

Merge dataframes and quick clean up

```{r}
od <- od_jo %>%
  bind_rows(od_kd)

# remove the sample that was taken after inflow pumps in dd6
od <- od %>%
  filter(time == "b") %>%
  select(-c(time))

# remove old dataframes
rm(list = c("od_jo", "od_kd"))   
```

Blanking (across each combination of experiment replicate and day)

```{r}
# group the dataframe by the unique combination of experiment_replicate and day
od <- od %>%
  group_by(experiment_replicate, day) %>%

  # create a new variable 'OD_control_value' that stores the OD value where OD_control == 1
  mutate(OD_control_value = ifelse(OD_control == 1, OD, NA)) %>%

  # fill missing values (NA) within each group by carrying the OD_control_value down and up
  fill(OD_control_value, .direction = "downup") %>%
  
  # create new varaible for blanked od
  mutate(od_blanked = OD - OD_control_value) %>%
  
  # set negative values to 0 (when controls were blanked with media - tiny differences)
  mutate(od_blanked = ifelse(od_blanked < 0, 0, od_blanked))
```

Some final cleaning 

```{r}
od <- od %>%
  
  # keep the reactor values only 
  filter(ID %in% c("M0", "M1", "M2", "M3",
                   "M4", "M5", "M6", "M7")) %>%
  
  # select the variables we want
  select(c(experiment_replicate, day, ID, od_blanked))
  
```


### Merge the three datasets and do some cleaning

```{r}
# merge all three dataframes
chemo_dd <- cfu %>%
  left_join(od) %>%
  left_join(outflow)

# create a factor for the day by experiment interaction
chemo_dd$sample <- interaction(chemo_dd$experiment_replicate, chemo_dd$day)

# remove controls and reactors that failed 
chemo_dd <- chemo_dd %>%
  
  # for dd2 none are removed (control was innoculated)
  
  # for dd3, remove control (M4)
  filter(!(ID %in% c("M4") & experiment_replicate == "dd3")) %>%
  
  # for dd4, remove control (M6)
  filter(!(ID %in% c("M6") & experiment_replicate == "dd4")) %>%

  # for dd6, remove control (M5) and M1, whose pump stopped during day 1
  filter(!(ID %in% c("M1", "M5") & experiment_replicate == "dd6")) %>%
  
  # for dd7, remove control (M2) and remove M3 from day 3, pump failed at end of day 2
  filter(!(ID %in% c("M2") & experiment_replicate == "dd7")) %>%
  filter(!(ID %in% c("M3") & experiment_replicate == "dd7" & day %in% c(3, 4)))

```


### Quick plots to compare across days within replicates

```{r}
ggplot(chemo_dd, aes(x = cfu_ul, y = dilution_rate, color = as.factor(day))) +
  geom_point() + 
  labs(x = "CFU/μL", y = "Dilution rate (ml/hour/ml)", color = "day") +
  theme_light() +
  facet_wrap(~ experiment_replicate, scales = "free_x")
```
```{r}
ggplot(chemo_dd, aes(x = od_blanked, y = dilution_rate, color = as.factor(day))) +
  geom_point() + 
  labs(x = "OD", y = "Dilution rate (ml/hour/ml)", color = "day") +
  theme_light() +
  facet_wrap(~ experiment_replicate, scales = "free_x")
```

### Plotting populations at equilibrium 

Ad hoc selection of sampling days based on equilibrium estimates from OD tracking and from sampling 

```{r}
chemo_dd_eqs <- chemo_dd %>%
  
  # overshoot only seen in the lowest dilution: take at least day 4
  filter(!(dilution_rate < 0.1 & day %in% c(1, 2, 3))) %>%
  
  # highest dilutions need time to get to equilibrium: take at least day 3 
  filter(!(dilution_rate > 0.7 & day %in% c(1, 2))) %>%
  
  # mid dilution rates get to equilibrium fast but could then evolve, take days 1 and 2 only
  filter(!(dilution_rate < 0.7 & dilution_rate > 0.1 & day %in% c(3, 4, 5, 6))) %>%
  
  # longer than 4 days is too much, equilibrium already reached but high risk of evolution
  filter(!(day > 4))

  
# Average across multiple relevant days 
# Mid dilutions, averaged across days 1 and 2
# High dilutions, averaged across days 3 and 4 
chemo_dd_eqs_av <- chemo_dd_eqs %>%
  select(-c(sample)) %>%
  group_by(experiment_replicate, ID, inflow_setting, 
           outflow_per_hour_mean, dilution_rate) %>%
  summarise(avg_od_blanked = mean(od_blanked, na.rm = TRUE),
            avg_cfu_ul = mean(cfu_ul, na.rm = TRUE))


```


```{r, fig.height=4, fig.width=7}

ggplot(chemo_dd_eqs_av, aes(x = avg_cfu_ul, y = dilution_rate, 
                            color = experiment_replicate)) +
  geom_point() + 
  labs(x = "CFU/μL", y = "Dilution rate (ml/hour/ml)", 
       color = "experiment") +
  theme_minimal()

```

```{r, fig.height=4, fig.width=7}

ggplot(chemo_dd_eqs_av, aes(x = avg_od_blanked, y = dilution_rate, 
                            color = experiment_replicate)) +
  geom_point() + 
  labs(x = "OD", y = "Dilution rate (ml/hour/ml)", 
       color = "experiment") +
  theme_minimal()

```




Quick plots of CFUs by dilution rate for each relevant day

```{r, fig.height=4, fig.width=7}

ggplot(chemo_dd_eqs, aes(x = cfu_ul, y = dilution_rate, 
                         color = experiment_replicate,
                         shape = as.factor(day))) +
  geom_point() + 
  labs(x = "CFU/μL", y = "Dilution rate (ml/hour/ml)", 
       color = "experiment",
       shape = "day") +
  theme_minimal()

```
Quick plots of od by dilution rate for each relevant day

```{r, fig.height=4, fig.width=7}
ggplot(chemo_dd_eqs, aes(x = od_blanked, y = dilution_rate, 
                         color = experiment_replicate,
                         shape = as.factor(day))) +
  geom_point() + 
  labs(x = "OD", y = "Dilution rate (ml/hour/ml)", 
       color = "experiment",
       shape = "day") +
  theme_minimal()
```


