---
title: "chemo-dd"
subtitle: "Supplementary plots"
output: 
  html_notebook:
    toc: true
    toc_depth: 4
author: "James Orr"
---

This notebook uses the processed data from `1-data-prep.Rmd` to create supplementary plots for our study on the shape of density dependence of *E. coli*. These plots are primarly for cross-referencing to ensure that our methods are appropriate and are robust to different data filtering options. Some plots will be in the supporting information accompanying the main article. 

This notebook produces:

- Plots of densities (OD and CFU) for all samples in each replicate as a function of dilution rate. (Supplementary Figure 1A).

- Plots of only equilibrium densities (OD and CFU) as a function of dilution rate for each data filtering option.

- Plots of the OD time series produced by the Chi.Bio reactors. These OD measurements are less precise than the daily OD readings that were done on plate readers. (Supplemenary Figure 1B)


### Set up environment 

Load packages and clear environment 

```{r}
#### Required packages
library(tidyverse) 
library(cowplot)

#### Clear  environment 
rm(list = ls())   
```

Load data

```{r}
all_samples <- read.csv("data/processed/all_samples.csv")
equilibrium_samples <- read.csv("data/processed/equilibrium_samples.csv")
chibio <- read.csv("data/processed/chibio.csv")

# "final" datasets
chemo_dd_eqs_av <- read.csv("data/processed/equilibrium_samples_av.csv")
chemo_dd_eqs_last <- read.csv("data/processed/equilibrium_samples_last.csv")
chemo_dd_eqs_first <- read.csv("data/processed/equilibrium_samples_first.csv")
chemo_dd_eqs_day1 <- read.csv("data/processed/equilibrium_samples_day1.csv")
chemo_dd_eqs_day2 <- read.csv("data/processed/equilibrium_samples_day2.csv")
chemo_dd_eqs_day3 <- read.csv("data/processed/equilibrium_samples_day3.csv")
chemo_dd_eqs_day4 <- read.csv("data/processed/equilibrium_samples_day4.csv")
```

### Densities per day for each replicate

```{r}
ggplot(all_samples, aes(y = cfu_ul, x = dilution_rate, color = day)) +
  geom_point() + 
  labs(y = "CFU/μL", x = "Dilution rate (ml/hour/ml)", color = "day") +
  scale_color_gradient(low = "#3BDCB4", high = "#185266", name = "Day") + 
  theme_light() +
  facet_wrap(~ experiment_replicate, scales = "fixed")
```
```{r}
ggplot(all_samples, aes(y = od_blanked, x = dilution_rate, color = day)) +
  geom_point() + 
  labs(y = "OD", x = "Dilution rate (ml/hour/ml)", color = "day") +
  scale_color_gradient(low = "#3BDCB4", high = "#185266", name = "Day") + 
  theme_light() +
  facet_wrap(~ experiment_replicate, scales = "fixed")
```

Cutoff between low, intermediate, and high dilution rates.

```{r, fig.height=6, fig.width=8}

ggplot(subset(all_samples, all_samples$day < 5), 
       aes(y = od_blanked, x = dilution_rate, 
           color = day, shape = experiment_replicate)) +
  
  # Add shaded regions
  geom_rect(aes(xmin = -Inf, xmax = 0.1, ymin = -Inf, ymax = Inf), 
            fill = "gray99", inherit.aes = FALSE) + 
  geom_rect(aes(xmin = 0.1, xmax = 0.65, ymin = -Inf, ymax = Inf), 
            fill = "gray95", inherit.aes = FALSE) + 
  geom_rect(aes(xmin = 0.65, xmax = Inf, ymin = -Inf, ymax = Inf), 
            fill = "gray90", inherit.aes = FALSE) +   
  
  geom_point(size = 4, alpha = 1) + 

  geom_vline(xintercept = c(0.1, 0.65), linetype = "dashed", color = "gray50") + # Vertical lines

  geom_line(aes(group = dilution_rate), linetype = "solid", 
            color = "black", size = 1.5, alpha = 0.3) + 

  scale_color_gradient(low = "#3BDCB4", high = "#185266", name = "Day") + 
  
  scale_shape_manual(values = c(15:19), name = "Replicate") +

  labs(y = "OD", x = "Dilution rate (ml/hour/ml)") + 
    
  theme_light()

```

Key observations from this plot are that: (i) in the "low" dilution rate zone (light grey) OD **decreases** each day and takes about 4 days to reach equilibrium due to the overshoot dynamic, (ii) in the "intermediate" dilution rate zone (medium grey) OD can go **up or down** (depending on stochasticity) so is considered at equilibrium on days 1 and 2 before evolution can have an effect, and (iii) in the "high dilution rate" zone (dark grey) OD **increases** each day until day 4 so is considered at equilibrium on day 4 only. 


Supplementary Figure 1A

```{r fig.height=8, fig.width=8}

ggplot(subset(all_samples, all_samples$day < 5), 
       aes(y = od_blanked, x = dilution_rate, 
           fill = as.factor(day), shape = experiment_replicate)) +
  
  geom_line(aes(group = dilution_rate), linetype = "solid", 
            color = "black", size = 1.5, alpha = 0.3) + 
  
  geom_point(size = 4, alpha = 0.7, colour = "gray50") + 
  
  # discrete colour scale
  scale_fill_manual(
    values = c(
      "1" = "#e5f5e0",
      "2" = "#a1d99b",
      "3" = "#41ab5d",
      "4" = "#005a25"),
      name = "Day") +
  
  scale_shape_manual(values = c(21:25), name = "Replicate") +

  # add in points from chemo_dd_eqs_av from main analysis
  #geom_point(data = equilibrium_samples, aes(y = od_blanked, 
  #                                       x = dilution_rate,
  #                                       #fill = day,
  #                                       shape = experiment_replicate),
  #           color = "red", size = 4, inherit.aes = FALSE, show.legend = FALSE) +
             
  labs(y = "OD", x = "Dilution rate (ml/hour/ml)") + 
    
  theme_minimal(base_size = 20) +
  
  # move legend into top right and make smaller and put the two side by side
  theme(legend.position = c(0.83, 0.87), legend.title = element_text(size = 12), 
        legend.text = element_text(size = 12),
        legend.box = "horizontal") +
  
  # tell legend to use the same shape and size for all points in legend
  guides(fill = guide_legend(override.aes = list(shape = 21, size = 5, colour = "gray30"))) +
  
  
  theme(panel.grid = element_blank(),
            panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5))


```


### Equilibrium densities by dilution rate

#### CFUs 

```{r, fig.height=5, fig.width=15}
# Define a function to generate the plot
make_plot <- function(data, title) {
  ggplot(data, aes(y = cfu_ul, 
                   x = dilution_rate, 
                   shape = experiment_replicate)) +
    geom_point(size = 2) + 
    labs(y = "CFU/μL", x = "Dilution rate (ml/hour/ml)",
         title = title) +
    scale_shape_manual(values = c(15:19), name = "Replicate") +
    theme_minimal()
}

# List of data frames
data_list <- list(
  equilibrium_samples, chemo_dd_eqs_av, chemo_dd_eqs_first, chemo_dd_eqs_last,
  chemo_dd_eqs_day1, chemo_dd_eqs_day2, chemo_dd_eqs_day3, chemo_dd_eqs_day4
)

# Corresponding plot titles
titles <- c(
  "all equilibrium", "equilibrium (mean)", 
  "equilibrium (first)", "equilibrium (last)",
  "day 1s", "day 2", "day 3", "day 4"
)


# Use Map to generate all plots
plots <- Map(make_plot, data_list, titles)

# Arrange them into a grid
plot_grid(plotlist = plots, nrow = 2)

rm(titles, data_list, plots, make_plot)

```

#### OD

```{r, fig.height=5, fig.width=15}

# Define a function to generate the plot
make_plot <- function(data, title) {
  ggplot(data, aes(y = od_blanked, 
                   x = dilution_rate, 
                   shape = experiment_replicate)) +
    geom_point(size = 2) + 
    labs(y = "OD", x = "Dilution rate (ml/hour/ml)",
         title = title) +
    scale_shape_manual(values = c(15:19), name = "Replicate") +
    theme_minimal()
}

# List of data frames
data_list <- list(
  equilibrium_samples, chemo_dd_eqs_av, chemo_dd_eqs_first, chemo_dd_eqs_last,
  chemo_dd_eqs_day1, chemo_dd_eqs_day2, chemo_dd_eqs_day3, chemo_dd_eqs_day4
)

# Corresponding plot titles
titles <- c(
  "all equilibrium", "equilibrium (mean)", 
  "equilibrium (first)", "equilibrium (last)",
  "day 1s", "day 2", "day 3", "day 4"
)


# Use Map to generate all plots
plots <- Map(make_plot, data_list, titles)

# Arrange them into a grid
plot_grid(plotlist = plots, nrow = 2)

rm(titles, data_list, plots, make_plot)

```

### Chi.Bio OD timeseries


```{r, fig.height=6, fig.width=10}
  
chibio_plot <- chibio %>%
  # remove control
  filter(!(ID %in% c("M2"))) %>%
  
  # remove M3 after day 2 (pump failed)
  filter(!(ID %in% c("M3") & hours > 48)) %>%
  
  # remove all points after day 4 (system stopped) 
  filter(!(hours > 96))
  

ggplot(chibio_plot, aes(x = hours, y = od_measured, 
                   group = ID, color = dilution_rate)) +
  
  # Each observation 
  geom_point(alpha = 0.1, size = 0.5) +
  
  # Smoothed lines for each ID
  geom_smooth(aes(group = ID), method = "loess", 
              se = FALSE, size = 1.2, span = 0.15) +
  
  # Set a continuous color gradient 
  scale_color_gradient(low = "#3BDCB4", high = "#185266") +
  
  # Add vertical lines for sampling moments
  geom_vline(xintercept = c(24, 48, 72, 96),
             linetype = "dashed", color = "grey40", size = 0.5) +
  
  # Add labels and title
  labs(x = "Hours", y = "OD", color = "Dilution Rate") +
  
  theme_minimal() +
  theme(legend.position = "right")
```

Supplementary Figure 1B

Select one high (above detection limit), one medium, and one low dilution rate that we have a full uninterrupted time series for. DD7 was the only experimental replicate where time series were uninterrupted for the full days. In that replicate, M0 has the highest dilution that is above detection limit. M4 has the lowest dilution (methodological control). M6 was disrupted half way through (e.g., lifted up to look at). That leaves M1 and M7 to choose from to represent the medium dilution rates. They are very similar, but M7's dilution rate closer to the middle of the full range of dilutions.  


```{r, fig.height=8, fig.width=8}
# filter one high, one medium, and one low dilution rates
chibio_plot2 <- chibio_plot %>%
  filter(ID %in% c("M0","M7", "M4")) %>%
  # remove negatives 
  filter(od_measured > 0.001)


ggplot(chibio_plot2, aes(x = hours, y = od_measured, 
                   group = ID, color = dilution_rate)) +
  
  # Each observation 
  geom_point(alpha = 0.5, size = 0.25) +
  
  # Smoothed lines for each ID
  #geom_smooth(aes(group = ID), method = "loess", 
  #            se = FALSE, size = 1, alpha = 0.25, span = 0.1) +
  
  # Set a continuous color gradient 
  #scale_color_gradient(low = "#3BDCB4", high = "#185266") +
  
  # Add labels and title
  labs(x = "Hours", y = "OD", color = "Dilution Rate") +
  
  theme_minimal(base_size = 20) +
  

  scale_y_log10(
    breaks = c(0.001, 0.01, 0.1, 1),
    labels = c("0.001", "0.01", "0.1", "1"),
    limits = c(0.001, 3)
  ) +
  
  # move legend into top right and make smaller and put the two side by side
  theme(legend.position = c(0.9, 0.15), legend.title = element_text(size = 12), 
        legend.text = element_text(size = 12),
        legend.box = "horizontal") +
  
  theme(panel.grid = element_blank(),
            panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5))

```
```{r}
unique(chibio_plot2$dilution_rate)
```




