---
title: "chemo-dd"
subtitle: "Supplementary plots"
output: 
  html_notebook:
    toc: true
    toc_depth: 4
author: "James Orr"
---

This notebook uses the processed data from `1-data-prep.Rmd` to create supplementary plots for our study on the shape of density dependence of *E. coli*. Here we plot: 

- densities (OD and CFU) for each experiment against dilution rate (colored by day)

- equilibrium density against dilution rate for all data to illustrate chosen cut-off points (low, intermediate, and high dilution rates), which determines what days a population was considered at equilibrium and unaffected by evolution. 

- equilibrium density against dilution rate for each data filtering option (for OD and CFU). 

- the chibio's OD timeseries for dd7. 


### Set up environment 

Load packages and clear environment 

```{r}
#### Required packages
library(tidyverse) 
library(cowplot)

#### Clear  environment 
rm(list = ls())   
```

Load data

```{r}
all_samples <- read.csv("data/processed/all_samples.csv")
equilibrium_samples <- read.csv("data/processed/equilibrium_samples.csv")
chibio <- read.csv("data/processed/chibio.csv")

# "final" datasets
chemo_dd_eqs_av <- read.csv("data/processed/equilibrium_samples_av.csv")
chemo_dd_eqs_last <- read.csv("data/processed/equilibrium_samples_last.csv")
chemo_dd_eqs_first <- read.csv("data/processed/equilibrium_samples_first.csv")
chemo_dd_eqs_day1 <- read.csv("data/processed/equilibrium_samples_day1.csv")
chemo_dd_eqs_day2 <- read.csv("data/processed/equilibrium_samples_day2.csv")
chemo_dd_eqs_day3 <- read.csv("data/processed/equilibrium_samples_day3.csv")
chemo_dd_eqs_day4 <- read.csv("data/processed/equilibrium_samples_day4.csv")
```

### Densities per day for each experiment

```{r}
ggplot(all_samples, aes(y = cfu_ul, x = dilution_rate, color = day)) +
  geom_point() + 
  labs(y = "CFU/μL", x = "Dilution rate (ml/hour/ml)", color = "day") +
  scale_color_gradient(low = "#3BDCB4", high = "#185266", name = "Day") + 
  theme_light() +
  facet_wrap(~ experiment_replicate, scales = "free_y")
```
```{r}
ggplot(all_samples, aes(y = od_blanked, x = dilution_rate, color = day)) +
  geom_point() + 
  labs(y = "OD", x = "Dilution rate (ml/hour/ml)", color = "day") +
  scale_color_gradient(low = "#3BDCB4", high = "#185266", name = "Day") + 
  theme_light() +
  facet_wrap(~ experiment_replicate, scales = "free_y")
```

Cutoff between low, intermediate, and high dilution rates. Only include up to and including day 4

```{r, fig.height=4, fig.width=8}

ggplot(subset(all_samples, all_samples$day < 5), 
       aes(y = od_blanked, x = dilution_rate, 
           color = day, shape = experiment_replicate)) +
  
  # Add shaded regions
  geom_rect(aes(xmin = -Inf, xmax = 0.1, ymin = -Inf, ymax = Inf), 
            fill = "gray99", inherit.aes = FALSE) + 
  geom_rect(aes(xmin = 0.1, xmax = 0.7, ymin = -Inf, ymax = Inf), 
            fill = "gray95", inherit.aes = FALSE) + 
  geom_rect(aes(xmin = 0.7, xmax = Inf, ymin = -Inf, ymax = Inf), 
            fill = "gray90", inherit.aes = FALSE) +   
  
  geom_point(size = 4, alpha = 1) + 

  geom_vline(xintercept = c(0.1, 0.7), linetype = "dashed", color = "gray50") + # Vertical lines

  geom_line(aes(group = dilution_rate), linetype = "solid", 
            color = "black", size = 1.5, alpha = 0.3) + 

  scale_color_gradient(low = "#3BDCB4", high = "#185266", name = "Day") + 
  
  scale_shape_manual(values = c(15:19), name = "Replicate") +

  labs(y = "OD", x = "Dilution rate (ml/hour/ml)") + 
    
  theme_light()


```

Key observations from this plot are that: (i) in the "low" dilution rate zone (light grey) OD **decreases** each day and takes about 4 days to reach equilibrium due to the overshoot dynamic, (ii) in the "intermediate" dilution rate zone (medium grey) OD can go **up or down** (depending on stochasticity) so is considered at equilibrium on days 1 and 2 before evolution can have an effect, and (iii) in the "high dilution rate" zone (dark grey) OD **increases** each day until day 4 so is considered at equilibrium on day 4 only. 


### Equilibrium densities by dilution rate

#### CFUs 

```{r, fig.height=5, fig.width=15}
# Define a function to generate the plot
make_plot <- function(data, title) {
  ggplot(data, aes(y = cfu_ul, 
                   x = dilution_rate, 
                   shape = experiment_replicate)) +
    geom_point(size = 2) + 
    labs(y = "CFU/μL", x = "Dilution rate (ml/hour/ml)",
         title = title) +
    scale_shape_manual(values = c(15:19), name = "Replicate") +
    theme_minimal()
}

# List of data frames
data_list <- list(
  equilibrium_samples, chemo_dd_eqs_av, chemo_dd_eqs_first, chemo_dd_eqs_last,
  chemo_dd_eqs_day1, chemo_dd_eqs_day2, chemo_dd_eqs_day3, chemo_dd_eqs_day4
)

# Corresponding plot titles
titles <- c(
  "all equilibrium", "equilibrium (mean)", 
  "equilibrium (first)", "equilibrium (last)",
  "day 1s", "day 2", "day 3", "day 4"
)


# Use Map to generate all plots
plots <- Map(make_plot, data_list, titles)

# Arrange them into a grid
plot_grid(plotlist = plots, nrow = 2)

rm(titles, data_list, plots, make_plot)

```

#### OD

```{r, fig.height=5, fig.width=15}

# Define a function to generate the plot
make_plot <- function(data, title) {
  ggplot(data, aes(y = od_blanked, 
                   x = dilution_rate, 
                   shape = experiment_replicate)) +
    geom_point(size = 2) + 
    labs(y = "OD", x = "Dilution rate (ml/hour/ml)",
         title = title) +
    scale_shape_manual(values = c(15:19), name = "Replicate") +
    theme_minimal()
}

# List of data frames
data_list <- list(
  equilibrium_samples, chemo_dd_eqs_av, chemo_dd_eqs_first, chemo_dd_eqs_last,
  chemo_dd_eqs_day1, chemo_dd_eqs_day2, chemo_dd_eqs_day3, chemo_dd_eqs_day4
)

# Corresponding plot titles
titles <- c(
  "all equilibrium", "equilibrium (mean)", 
  "equilibrium (first)", "equilibrium (last)",
  "day 1s", "day 2", "day 3", "day 4"
)


# Use Map to generate all plots
plots <- Map(make_plot, data_list, titles)

# Arrange them into a grid
plot_grid(plotlist = plots, nrow = 2)

rm(titles, data_list, plots, make_plot)

```


### Chibio OD timeseries


```{r, fig.height=6, fig.width=10}
  
chibio_plot <- chibio %>%
  # remove control
  filter(!(ID %in% c("M2"))) %>%
  
  # remove M3 after day 2 (pump failed)
  filter(!(ID %in% c("M3") & hours > 48)) %>%
  
  # remove all points after day 4 (system stopped) 
  filter(!(hours > 96))
  

ggplot(chibio_plot, aes(x = hours, y = od_measured, 
                   group = ID, color = dilution_rate)) +
  
  # Each observation 
  geom_point(alpha = 0.1, size = 0.5) +
  
  # Smoothed lines for each ID
  geom_smooth(aes(group = ID), method = "loess", 
              se = FALSE, size = 1.2, span = 0.15) +
  
  # Set a continuous color gradient 
  scale_color_gradient(low = "#3BDCB4", high = "#185266") +
  
  # Add vertical lines for sampling moments
  geom_vline(xintercept = c(24, 48, 72, 96),
             linetype = "dashed", color = "grey40", size = 0.5) +
  
  # Add labels and title
  labs(x = "Hours", y = "OD", color = "Dilution Rate") +
  
  theme_minimal() +
  theme(legend.position = "right")
```


