---
title: "chemo-dd"
subtitle: "Supplementary models"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
author: "James Orr"
---

This notebook uses the processed data from `1-data-prep.Rmd` to create supplementary models for our study on the shape of density dependence of *E. coli*. Our goal was to using a Bayesian regression framework to fit non-linear models to the empirical data to estimate the shape of density dependence. 

We considered an **inverted theta-logistic model** and a simple **power model**. 


### Set up environment 

Load packages and clear environment 

```{r}
#### Required packages
library(tidyverse)  
library(brms)
library(cowplot)

#### Clear  environment 
rm(list = ls())   
```

```{r}
all_samples <- read.csv("data/processed/all_samples.csv")
equilibrium_samples <- read.csv("data/processed/equilibrium_samples.csv")
chibio <- read.csv("data/processed/chibio.csv")

# "final" datasets
chemo_dd_eqs_av <- read.csv("data/processed/equilibrium_samples_av.csv")
chemo_dd_eqs_last <- read.csv("data/processed/equilibrium_samples_last.csv")
chemo_dd_eqs_first <- read.csv("data/processed/equilibrium_samples_first.csv")
chemo_dd_eqs_day1 <- read.csv("data/processed/equilibrium_samples_day1.csv")
chemo_dd_eqs_day2 <- read.csv("data/processed/equilibrium_samples_day2.csv")
chemo_dd_eqs_day3 <- read.csv("data/processed/equilibrium_samples_day3.csv")
chemo_dd_eqs_day4 <- read.csv("data/processed/equilibrium_samples_day4.csv")

#theory_predictions <- read.csv("data/theory/prediction.csv")
#theory_predictions2 <- read.csv("data/theory/prediction2.csv")

```

To begin with we'll use `chemo_dd_eqs_av`, which we believe is the most sensible way of selecting equilibrium points. This dataset contains 31 observations and has no pseudoreplication.

### Flipped theta-logistic model 

We manipulated per capita growth through dilution rate and equilibrium density measured by OD is our response variable. This means we need to fit an inverted theta-logistic model.

The standard theta-logistic model is: 

$$
\frac{1}{N} \frac{d N}{d t}=r\left(1-\left(\frac{N}{K}\right)^\theta\right)
$$
Where $\frac{1}{N} \frac{d N}{d t} = g$ is the per capita growth (i.e. `dilution_rate`), $N$ is the density (i.e., `od_blanked`), $r$ is the intrinsic growth rate, $K$ is the carrying capacity, and $\theta$ is the parameter that controls the shape of density dependence. When $\theta$ is 1, it is linear (i.e., logistic growth model), when it is >1 the density dependence is concave (what we expect based on consumer-resource theory), when it is <1 the density dependence is sublinear. 

We will be using `brms` to estimate the values of $r$, $K$, and most importantly $\theta$, based on the relationship between $g$ and $N$. However, as we manipulated per capita growth, we need to express the theta-logistic model in terms of $N$ (our response variable given our experimental design). 

Starting from

$$
g=r\left(1-\left(\frac{N}{K}\right)^\theta\right)
$$
we can rearrange to

$$
N=K\left(1-\frac{g}{r}\right)^{\frac{1}{\theta}}
$$
The data we're trying to fit the model to are: 

```{r, fig.height=3, fig.width=4}
ggplot(chemo_dd_eqs_av, aes(y = od_blanked, 
                                      x = dilution_rate, 
                                      shape = experiment_replicate)) +
  geom_point(size = 2) + 
  labs(y = "OD", x = "Dilution rate (ml/hour/ml)",
       title = "chemo_dd_eqs_av") +
  scale_shape_manual(values = c(15:19), name = "Replicate") +
  theme_minimal()
```
We'll use Gamma distribution with the identity link to account for the fact that we can't have negative OD values. 

**Priors** 

$\theta$ describes the curvature of the model, $r$ describes the intercept with the x axis (maximum growth rate), $K$ describes the intercept with the y axis (carrying capacity). Based on prior knowledge, we can provide some biologically informed priors for growth rates and carrying capacity based on previous experiments with this system. Based on pilot studies and related projects, we expected $r$ to be around 1 and $K$ to be around 0.2. However, when fitting the model we will use much looser priors to show that our results are not influenced by the choice of priors. 

- We'll use a prior for $r$ that is uniform between 0.5 and 1.5 

- We'll use a prior for $K$ that is between 0.1 and 0.5

With our bacteria growing in chemostats supplied with 0.05% glucose M9 the extreme values of these priors are extremely unlikely, but we want to be conservative and not bias our results by using priors that are too tight.

We'll give extremely loose priors for $\theta$, as this is the main parameter we're interest in, and we don't want to potentially bias the results. So we'll use a uniform prior between 0.1 (highly sublinear) and 10 (highly superlinear). As theta is an exponent, 10 and -1 are symmetrical around 1.


```{r, include = F}

model <- brm(
  
  # model formula (parameters to fit are identified and nl flag is set)
  bf(od_blanked ~ K * (1 - (dilution_rate / r)) ^ (1 / theta),
     K ~ 1,
     r ~ 1, 
     theta ~ 1,
     nl = TRUE
  ),
  
  # data to use
  data = chemo_dd_eqs_av,
  
  # errors are gamma with identity link function (OD can't be negative)
  family = Gamma(link = "identity"), 
  
  # priors 
  prior = c(
    prior(uniform(0.5, 1.5), nlpar = "r", lb = 0.5, ub = 1.5),  
    prior(uniform(0.1, 0.5), nlpar = "K", lb = 0.1, ub = 0.5), 
    prior(uniform(0.1, 10), nlpar = "theta", lb = 0.1, ub = 10)  
  ),
  
  # hyperparameters 
  iter = 3000, warmup = 1500, chains = 4,
  control = list(adapt_delta = 0.9,
                 max_treedepth = 10),
  
)
```

```{r}
summary(model)
plot(model)
conditional_effects(model)
pairs(model)
pp_check(model, type = "scatter_avg")
```

```{r}
# Extract predictions from the model
conditional_effects_data <- conditional_effects(model)
predicted_data <- as.data.frame(conditional_effects_data$dilution_rate)
posteriors <- posterior_samples(model)
K <- mean(posteriors$`b_K_Intercept`)
r <- mean(posteriors$`b_r_Intercept`)            # use median for skewed posteriors? 
theta <- mean(posteriors$`b_theta_Intercept`)


# curve prediction with proper theta logistic model with estimated parameters 
od_seq <- seq(0, 0.2, by = 0.001)
dilution_curve <- r * (1 - (od_seq / K)^ theta)
curve_data2 <- data.frame(od_blanked = od_seq, 
                         dilution_rate = dilution_curve)
curve_data2 <- curve_data2 %>% filter(dilution_rate > 0)


# Create the plot with the original data points and model predictions
p1 <- ggplot(chemo_dd_eqs_av, aes(y = od_blanked, x = dilution_rate)) +
  geom_point() +
  labs(y = "OD", x = "Dilution rate (ml/hour/ml)", 
       title = "Growth-Density inversion") +
  # brms model predictions
  geom_line(data = predicted_data, aes(y = estimate__, 
                                       x = dilution_rate)) +  
  geom_ribbon(data = predicted_data, aes(ymin = lower__, 
                                         ymax = upper__, 
                                         x = dilution_rate), 
              alpha = 0.2, inherit.aes = FALSE) + 

  theme_minimal()


# Density-dependent plot
p2 <- ggplot(chemo_dd_eqs_av, aes(x = od_blanked, y = dilution_rate)) +
  geom_point() +
  labs(x = "OD", y = "Dilution rate (ml/hour/ml)", 
       title = "Density dependence") +
  # theta-logistic
  geom_line(data = curve_data2, aes(x = od_blanked, y = dilution_rate), 
              color = "black", size = 1) +
  theme_minimal()


plot_grid(p1, p2)

```

Black line is the model with the gamma errors. The $r$ bounding is causing fitting problems. If $r$ was able to be lower, then $\theta$ would probably be higher. The two parameters are correlated in the model, and the bounding of $r$ is likely causing a bias in the estimate of $\theta$.

Despite this bias to lower values of $\theta$, $\theta$ is still clearly above one (mean of 1.76, error of 0.28, l-95% CI of 1.25, and u-95% CI or 2.37). 


### Power model 

As the inverse-theta logistic model is tricky to fit, we can use another model that can quantify the shape of the curvature. A simple power model that has a single parameter that describes its curvature is: 

$$ y = ax^\frac{1}{\beta} + c $$
Where $y$ is equal to OD, $x$ is equal to dilution rate, $a$ is related to where the curve cross the $x$ axis (but not quite intercept as it varies depending on $\beta$ (it must be negative for the function to be decreasing), $\beta$ is the curvature (linear when 1, sublinear when greater than 1, and superlinear when lower than 1, must be above 0), and $c$ is the y intercept (when x is 0). 

We can fit this with very loose priors (that are hard to intepret biologically).

```{r, include = F}

power_model <- brm(
  # Model formula (parameters to fit are identified and nl flag is set)
  bf(od_blanked ~ a * (dilution_rate^(1/b)) + c, 
    a ~ 1,  
    b ~ 1,  
    c ~ 1,  
    nl = TRUE
  ),
  # Data to use
  data = chemo_dd_eqs_av,
  
  # Errors are gamma with identity link function
  family = Gamma(link = "identity"),
  
  # Priors
  prior = c(
    prior(uniform(-5, 5), nlpar = "a", lb = -5, ub = 5),
    prior(uniform(0, 2), nlpar = "b", lb = 0, ub = 2),
    prior(uniform(0, 2), nlpar = "c", lb = 0, ub = 2)
  ),
  
  # Hyperparameters
  iter = 3000, 
  warmup = 1500, 
  chains = 4,
  control = list(adapt_delta = 0.9, max_treedepth = 10)
)

```

```{r}
summary(power_model)
plot(power_model)
conditional_effects(power_model)
pairs(power_model)
pp_check(power_model, type = "scatter_avg")
```


```{r}
# Extract predictions from the model
conditional_effects_data <- conditional_effects(power_model)
predicted_data <- as.data.frame(conditional_effects_data$dilution_rate)
posteriors <- posterior_samples(power_model)
a <- mean(posteriors$`b_a_Intercept`)
b <- mean(posteriors$`b_b_Intercept`)          
c <- mean(posteriors$`b_c_Intercept`)

# Flipped power model for density dependence with estimated parameters
od_seq <- seq(0, 0.25, by = 0.001)
dilution_curve <- ((od_seq - c) / a)^b
curve_data <- data.frame(od_blanked = od_seq, 
                         dilution_rate = dilution_curve)
curve_data <- curve_data %>% filter(dilution_rate > 0)


# Model predictions and empirical observations
p1 <- ggplot(chemo_dd_eqs_av, aes(y = od_blanked, x = dilution_rate)) +
  geom_point() +
  labs(y = "OD", x = "Dilution rate (ml/hour/ml)", 
       title = "Growth-Density inversion") +
  # brms model predictions
  geom_line(data = predicted_data, aes(y = estimate__, 
                                       x = dilution_rate)) +  
  geom_ribbon(data = predicted_data, aes(ymin = lower__, 
                                         ymax = upper__, 
                                         x = dilution_rate), 
              alpha = 0.2, inherit.aes = FALSE) + 
  theme_minimal()

# Density-dependent plot
p2 <- ggplot(chemo_dd_eqs_av, aes(x = od_blanked, y = dilution_rate)) +
  geom_point() +
  labs(x = "OD", y = "Dilution rate (ml/hour/ml)", 
       title = "Density dependence") +
  
  # theta-logistic
  geom_line(data = curve_data, aes(x = od_blanked, y = dilution_rate), 
              color = "black", size = 1) +
  
  theme_minimal()


plot_grid(p1, p2)

```


### Fit models to every subset of data

**Power model**

Create function that runs models and makes plots

```{r}
# Function to analyze dataset and save results as objects
analyze_dataset <- function(dataset_name) {
  # Load dataset dynamically
  dataset <- get(dataset_name)
  
  # Run the model
  power_model <- brm(
    bf(od_blanked ~ a * (dilution_rate^(1/b)) + c, 
       a ~ 1, b ~ 1, c ~ 1, nl = TRUE),
    data = dataset,
    family = Gamma(link = "identity"),
    prior = c(
      prior(uniform(-5, 5), nlpar = "a", lb = -5, ub = 5),
      prior(uniform(0, 2), nlpar = "b", lb = 0, ub = 2),
      prior(uniform(0, 2), nlpar = "c", lb = 0, ub = 2)
    ),
    iter = 3000, warmup = 1500, chains = 4,
    control = list(adapt_delta = 0.95, max_treedepth = 12)
  )
  
  # Save summary as an object
  model_summary <- summary(power_model)
  
  # Generate pp_check plot
  pp_plot <- pp_check(power_model, type = "scatter_avg")
  
  # Generate conditional effects plot
  conditional_effects_data <- conditional_effects(power_model)
  predicted_data <- as.data.frame(conditional_effects_data$dilution_rate)
  
  p1 <- ggplot(dataset, aes(y = od_blanked, x = dilution_rate)) +
    labs(
      y = "", x = "") +
  
    geom_ribbon(
      data = predicted_data,
      aes(ymin = lower__, ymax = upper__, x = dilution_rate),
      alpha = 0.4, inherit.aes = FALSE, fill = "#3BDCB4", color = NA
    ) +
    
    geom_line(data = predicted_data, 
              aes(y = estimate__, x = dilution_rate), 
              color = "#185266", size = 3) +
    
    geom_point(colour = "#185266", size = 5) +
    
    theme_light(base_size = 25) 
  
  ################### theta estimate ############################
  posterior <- posterior_samples(power_model)
  theta_samples <- posterior$`b_b_Intercept`
  theta_mean <- posterior_summary(power_model)["b_b_Intercept", 1]
  theta_low <- posterior_summary(power_model)["b_b_Intercept", 3]
  theta_high <- posterior_summary(power_model)["b_b_Intercept", 4]
  
  # Calculate density for posterior samples
  density_data <- density(theta_samples)
  density_df <- data.frame(
    theta = density_data$x,
    density = density_data$y) 
  
  theta_density_at_mean <- approx(density_data$x, density_data$y, xout = theta_mean)$y
  
  p2 <- ggplot(density_df, aes(x = theta, y = density)) +
    geom_line(color = "#185266", size = 2) +
    geom_area(
      data = density_df %>%
        filter(theta >= theta_low & theta <= theta_high),
      aes(x = theta, y = density),
      fill = "#3BDCB4", alpha = 0.4) +
    geom_segment(
      aes(x = theta_mean, xend = theta_mean, 
          y = 0, yend = theta_density_at_mean),
      color = "#185266", size = 2) +
    labs(
      x = "",
      y = ""
    ) +
    geom_vline(xintercept = 1, color = "gray80", size = 2, linetype = 2) +
    scale_x_log10(
      limits = c(0.15, 6.66),                # Set range of x-axis
      breaks = c(0.2, 0.5, 1, 2, 5)          # Logarithmic breaks
    ) +
    theme_light(base_size = 25) 
  
    
  # Return all results as a list
  return(list(
    summary = model_summary,
    pp_check_plot = pp_plot,
    fit = p1,
    posteriors = p2
  ))
}

```

Run the function for each dataset and store everything in a list

```{r, include=FALSE}
# List of dataset names
datasets <- c("chemo_dd_eqs_av", "chemo_dd_eqs_last", "chemo_dd_eqs_first", 
              "chemo_dd_eqs_day1", "chemo_dd_eqs_day2", "chemo_dd_eqs_day3", 
              "chemo_dd_eqs_day4")

# Run the function for all datasets and save results in a named list
results <- lapply(datasets, function(name) {
  analyze_dataset(name)
})
names(results) <- datasets

# Access example outputs:
# - Model summary for `chemo_dd_eqs_av`: results[[1]]$summary

```


```{r, fig.height=12, fig.width=36}

# Extract "fit" and "posteriors" plots for each dataset
fit_plots <- lapply(results, function(res) res$fit)
posteriors_plots <- lapply(results, function(res) res$posteriors)


# Create plot grids for the top (fit) and bottom (posteriors) rows
fit_row <- plot_grid(plotlist = fit_plots, ncol = 7, scale = 0.9,
                     labels = datasets, label_size = 16, label_x = 0.15)
posteriors_row <- plot_grid(plotlist = posteriors_plots, ncol = 7, scale = 0.9)

# Combine rows with titles
plot_grid(
  fit_row, posteriors_row, 
  ncol = 1
)
```

**Flipped theta logistic**

```{r}
# Function to analyze dataset and save results as objects
analyze_dataset2 <- function(dataset_name) {
  # Load dataset dynamically
  dataset <- get(dataset_name)
  
  # Run the model
  ft_model <- brm(
    bf(od_blanked ~ K * (1 - (dilution_rate / r)) ^ (1 / theta),
       K ~ 1, 
       r ~ 1, 
       theta ~ 1, 
       nl = TRUE
       ),
    data = dataset,
    family = Gamma(link = "identity"), 
    prior = c(
      prior(uniform(0.9, 1.2), nlpar = "r", lb = 0.9, ub = 1.2),  
      prior(uniform(0.10, 0.25), nlpar = "K", lb = 0.10, ub = 0.25), 
      prior(uniform(-0.5, 5), nlpar = "theta", lb = -0.5, ub = 5) 
      ),
    iter = 3000, warmup = 1500, chains = 4,
    control = list(adapt_delta = 0.9,
                   max_treedepth = 10),
    )

  
  # Save summary as an object
  model_summary <- summary(ft_model)
  
  # Generate pp_check plot
  pp_plot <- pp_check(ft_model, type = "scatter_avg")
  
  # Generate conditional effects plot
  conditional_effects_data <- conditional_effects(ft_model)
  predicted_data <- as.data.frame(conditional_effects_data$dilution_rate)
  
  p1 <- ggplot(dataset, aes(y = od_blanked, x = dilution_rate)) +
    labs(
      y = "", x = "") +
  
    geom_ribbon(
      data = predicted_data,
      aes(ymin = lower__, ymax = upper__, x = dilution_rate),
      alpha = 0.4, inherit.aes = FALSE, fill = "#3BDCB4", color = NA
    ) +
    
    geom_line(data = predicted_data, 
              aes(y = estimate__, x = dilution_rate), 
              color = "#185266", size = 3) +
    
    geom_point(colour = "#185266", size = 5) +
    
    theme_light(base_size = 25) 
  
  ################### theta estimate ############################
  posterior <- posterior_samples(ft_model)
  theta_samples <- posterior$`b_theta_Intercept`
  theta_mean <- posterior_summary(ft_model)["b_theta_Intercept", 1]
  theta_low <- posterior_summary(ft_model)["b_theta_Intercept", 3]
  theta_high <- posterior_summary(ft_model)["b_theta_Intercept", 4]
  
  # Calculate density for posterior samples
  density_data <- density(theta_samples)
  density_df <- data.frame(
    theta = density_data$x,
    density = density_data$y) 
  
  theta_density_at_mean <- approx(density_data$x, density_data$y, xout = theta_mean)$y
  
  p2 <- ggplot(density_df, aes(x = theta, y = density)) +
    geom_line(color = "#185266", size = 2) +
    geom_area(
      data = density_df %>%
        filter(theta >= theta_low & theta <= theta_high),
      aes(x = theta, y = density),
      fill = "#3BDCB4", alpha = 0.4) +
    geom_segment(
      aes(x = theta_mean, xend = theta_mean, 
          y = 0, yend = theta_density_at_mean),
      color = "#185266", size = 2) +
    labs(
      x = expression(theta),
      y = "Frequency"
    ) +
    geom_vline(xintercept = 1, color = "gray80", size = 2, linetype = 2) +
    scale_x_log10(
      limits = c(0.25, 4),                # Set range of x-axis
      breaks = c(0.25, 0.5, 1, 2, 4)     # Logarithmic breaks
    ) +
    theme_light(base_size = 25) 
  
    
  # Return all results as a list
  return(list(
    summary = model_summary,
    pp_check_plot = pp_plot,
    fit = p1,
    posteriors = p2
  ))
}

```

Run the function for each dataset and store everything in a list

```{r, include=FALSE}
# List of dataset names
datasets <- c("chemo_dd_eqs_av", "chemo_dd_eqs_last", "chemo_dd_eqs_first", 
              "chemo_dd_eqs_day1", "chemo_dd_eqs_day2", "chemo_dd_eqs_day3", 
              "chemo_dd_eqs_day4")

# Run the function for all datasets and save results in a named list
results <- lapply(datasets, function(name) {
  analyze_dataset2(name)
})
names(results) <- datasets

# Access example outputs:
# - Model summary for `chemo_dd_eqs_av`: results[[1]]$summary

```


```{r, fig.height=12, fig.width=36}

# Extract "fit" and "posteriors" plots for each dataset
fit_plots <- lapply(results, function(res) res$fit)
posteriors_plots <- lapply(results, function(res) res$posteriors)


# Create plot grids for the top (fit) and bottom (posteriors) rows
fit_row <- plot_grid(plotlist = fit_plots, ncol = 7, scale = 0.9,
                     labels = datasets, label_size = 16, label_x = 0.15)
posteriors_row <- plot_grid(plotlist = posteriors_plots, ncol = 7, scale = 0.9)

# Combine rows with titles
plot_grid(
  fit_row, posteriors_row, 
  ncol = 1
)
```

In summary, both the flipped theta logistic model and the power model show strong support for superlinear density dependence for all sensible filtering options. Defining equilibrium based on our heuristics makes most sense (columns one, two and three), and maybe using the final day of the experiment (day 4, last column) is also justifiable. For those four filtering options there is support for superlinear density dependence. When only day 1 or only day 2 data are used, the overshooting in low dilution rates is very extreme and this creates a poor fit where there is no statistical support for non-linear density dependence. 




